Drivers of spatial synchrony in giant kelp populations

### 2021-05-14 paper/SupMat.Rnw
changing paranthesis

\noindent for species $i=1,2$. Here, $N_i(t)$ is the population density of species $i$ at time $t$, and $N = N_1(t)+N_2(t)$ is fixed through time. The parameter, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundiity of specise $i$ at time $t$. The model postulates that individuals die at rate $\delta$ at each time step, and are replaced by juveniles in population to the reproductive output of the two species that year. We take, for each $i$, $B_i(t)$ to be $iid$ through time. We let $B_i = \exp(b_i)$, where $(b_1, b_2)$ is some distribution such that $b_i \sim N(\mu_i, \sigma^2)$ and such that $\CovX(b_1, b_2) = \rho$, a fixed quantity we will specify later. However, $(b_1, b_2)$ is not necessarily a bivariate normal distribution. We will consider various distributions of $(b_1, b_2)$ with these properties, corresponding to our symmetric and asymmetric tail associations cases defined in the main text and elaborated below. We assume without loss of generality that $\mu_1 \leq \mu_2$, so that species 1 is the weaker competitor.

@@

We created three cases for $(b_1,b_2)$, the case of symmetric tail association and the cases of asymmetrical left- and right-tail association. To generate $M$ points $(b_1^{(i)}, b_2^{(i)})$, $i=1,..,M$, for the left tail association case, we first generated M points $(a_1^{(i)},a_2^{(i)})$, $i=1,...,M$, from a bivariate normal distribution $N([0,0], \Sigma)$, where here $\Sigma = \begin{pmatrix} 1&0\\0&1\end{pmatrix}$. Then, for each index $i$, we randomly, with 50\% probability, either (A) let $\tilde b_1^{(i)} = -|a_1^{(i)}|$ and $\tilde b_2^{(i)} = -a|_1^{(i)}|$, or (B) let $\tilde b_1^{(i)} = |a_1^{(i)}|$ and $\tilde b_2^{(i)} = |a_2^{(i)}|$. We then let $b_1^{(i)} = \sigma \tilde b_1^{(i)} + \mu_1$ and $b_2^{(i)} = \sigma \tilde b_2^{(i)} + \mu_2$. Perfect association in the left tails of the resulting distribution $(b_1,b_2)$ results from the fact that, in case (A) above, both $\tilde b_1^{(i)}$ and $\tilde b_2^{(i)}$ were assigned to the same value. For speed of computation, when samples $(\tilde b_1^{(i)},\tilde b_2^{(i)})$, $i=1,...,M$ were needed for every distinct set of values of the parameters $\mu_1$, $\mu_2$, and $\sigma^2$, we used the same $(\tilde b_1^{(i)},\tilde b_2^{(i)})$, $i=1,...,M$, simply transforming these values differently (using the different valus of $\mu_1$, $\mu_2$, $\sigma$) for each set of parameter values. Right-tail associated noise was generated in an analogous way.

Noise with symmetric tail association was generated by taking a very large sample (100000) of left-tail associated noise, as described above, and then calculating the Pearson correaltion coefficient, rho. The value obtained was ... . The symmetric noise was then generated once from $N([0,0], \Sigma)$ where $\Sigma = \begin{pmatrix} 1&\rho \\ \rho&1\end{pmatrix}$, denoting the points so generated by $(\tilde b_1^{(i)},\tilde b_2^{(i)})$ for $i=1,...,M$. We then set $b_1^{(i)} = \sigma \tilde b_1^{(i)} + \mu_1$ and $b_2^{(i)} = \sigma \tilde b_2^{(i)}+ \mu_2$.

@@

\begin{equation}
\bar r_1^\#=\EX(\ln [(1-\delta)+\delta \exp(b_1^\#-b_2)]),  \label{rbar1sharp_E}
\end{equation}

@@

\begin{equation}
r_2=(\ln (\frac{N_2(t+1)}{N_2(t)}))=0, \label{r2_is_zero}
\end{equation}

@@

\begin{equation}
\Delta I=(\bar r_1 - \bar r_1^\#)-(\bar r_2- \bar r_2^\#)=\bar r_1 - \bar r_1^\#+\bar r_2^\#  \label{DeltaI}
\end{equation}

\noindent has been defined in terms of expected values of elementary expressions of the random variables $(b_1,b_2)$, $(b_1^\#,b_2)$, and $(b_2^\#, b_2)$.

Having already specified that $b_1$ (and therefore also $b_1^\#$) is distributed as $N(\mu_1, \sigma^2)$, and that $b_2$ (and therefore also $b_2^\#$) is distributed as $N(\mu_2, \sigma^2)$, we know, using $\Sigma$ as notation for a covariance matrix that

\begin{equation}
(b_1^\#, b_2) \sim N(\begin{pmatrix} \mu_1\\\mu_2\end{pmatrix}, \Sigma) \label{b1sharpb2_distribution}
\end{equation}

@@

\begin{equation}
(b_2^\#, b_2) \sim N(\begin{pmatrix} \mu_2\\\mu_2\end{pmatrix}, \Sigma) \label{b2sharpb2_distribution}
\end{equation}

@@

\begin{equation}
(b_1,b_2) \sim  N(\begin{pmatrix} \mu_1\\\mu_2\end{pmatrix}, \Sigma) \label{b1b2_distribution_sym}
\end{equation}

@@

\begin{equation}
\Sigma = \begin{pmatrix} \sigma^2&\rho\\\rho&\sigma^2\end{pmatrix}. \label{Sigmamat_rho}
\end{equation}

@@

\begin{enumerate}
\item Generate left-tail assocated noise $(\tilde b_{l1}^{(i)}, \tilde b_{l2}^{(i)})$, $i=1,...,M$, as described in the section \ref{sect:noise}. Hence $\tilde b_{l1}^{i)}$ and $\tilde b_{l2}^{(i)}$ are distributed as $N(0,1)$. 
\item Generate right-tail associated noise $(\tilde b_{r1}^{(i)}, \tilde b_{r2}^{(i)})$, $i=1,...,M$, via the analogous procedure. Again, $\tilde b_{r1}^{i)}$ and $\tilde b_{r2}^{(i)}$ are distributed as $N(0,1)$. 
\item Generate $M$ points $\tilde u_i$, $i = 1,...,M$ from a standard normal distribution. 
\end{enumerate}

@@

\item $\hat {\Delta I} = \hat \bar r_1 - \hat \bar r_1^{\#} + \hat \bar r_2^{\#}$, estimated ?syparatilly? for the left-tail associated, right-tail associated and symmetric noise cases. We obtained conservative standard error estimates by adding the standard error of the estimates of $\hat \bar r_1, \hat \bar r_1^{\#}$, and $\hat \bar r_2^{\#}$.

@@

Not only does the expected value $\bar r_1$ influence the capacity for species 1 to invade where rare, but so do other aspectf of the distribution of
\begin{equation}
r_1 = \ln[1-\delta+\delta \exp(b_1-b_2)]. \label{r1_distribution}
\end{equation}

@@

We also considered the probability $\Prob[r_1 >0]$, which is, explicitly, the probability that environmental variables that permit invasion. We have
\begin{align}
\Prob[r_1>0] &= \Prob [1-\delta+\delta \exp(b_1-b_2)] > 1] \\
&= \Prob[\exp (b_1-b_2)>1] \\
&= \Prob[b_1>b_2],
\end{align}

>>>>

\noindent for species $i=1,2$. 
Here, $N_i(t)$ is the population density of species $i$ at time $t$, and $N = N_1(t)+N_2(t)$ is fixed through time. 
The parameter, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundiity of specise $i$ at time $t$. 
The model postulates that individuals die at rate $\delta$ at each time step, and are replaced by juveniles in population to the reproductive output of the two species that year.
We take, for each $i$, $B_i(t)$ to be $iid$ through time. 
We let $B_i = \exp(b_i)$, where $\left(b_1, b_2\right)$ is some distribution such that $b_i \sim N\left(\mu_i, \sigma^2\right)$ and such that $\CovX(b_1, b_2) = \rho$, a fixed quantity we will specify later.
However, $\left(b_1, b_2\right)$ is not necessarily a bivariate normal distribution.
We will consider various distributions of $\left(b_1, b_2\right)$ with these properties, corresponding to our symmetric and asymmetric tail associations cases defined in the main text and elaborated below.
We assume without loss of generality that $\mu_1 \leq \mu_2$, so that species 1 is the weaker competitor.

@@

We created three cases for $(b_1,b_2)$, the case of symmetric tail association and the cases of asymmetrical left- and right-tail association. 
To generate $M$ points $\left(b_1^{(i)}, b_2^{(i)}\right)$, $i=1,..,M$, for the left tail association case, we first generated M points $(a_1^{(i)},a_2^{(i)})$, $i=1,...,M$, from a bivariate normal distribution $N\left( \vec{0}, \Sigma \right)$, where here $\Sigma = \begin{pmatrix} 1&0\\0&1\end{pmatrix}$. 
Then, for each index $i$, we randomly, with 50\% probability, either (A) let $\tilde b_1^{(i)} = -|a_1^{(i)}|$ and $\tilde b_2^{(i)} = -a|_1^{(i)}|$, or (B) let $\tilde b_1^{(i)} = |a_1^{(i)}|$ and $\tilde b_2^{(i)} = |a_2^{(i)}|$. 
We then let $b_1^{(i)} = \sigma \tilde b_1^{(i)} + \mu_1$ and $b_2^{(i)} = \sigma \tilde b_2^{(i)} + \mu_2$. 
Perfect association in the left tails of the resulting distribution $\left(b_1,b_2\right)$ results from the fact that, in case (A) above, both $\tilde b_1^{(i)}$ and $\tilde b_2^{(i)}$ were assigned to the same value. 
For speed of computation, when samples $(\tilde b_1^{(i)},\tilde b_2^{(i)})$, $i=1,...,M$ were needed for every distinct set of values of the parameters $\mu_1$, $\mu_2$, and $\sigma^2$, we used the same $\left(\tilde b_1^{(i)},\tilde b_2^{(i)}\right)$, $i=1,...,M$, simply transforming these values differently (using the different valus of $\mu_1$, $\mu_2$, $\sigma$) for each set of parameter values. Right-tail associated noise was generated in an analogous way.

Noise with symmetric tail association was generated by taking a very large sample (100000) of left-tail associated noise, as described above, and then calculating the Pearson correaltion coefficient, rho.
The value obtained was ... . 
The symmetric noise was then generated once from $N\left(\vec{0}, \Sigma \right)$ where $\Sigma = \begin{pmatrix} 1&\rho \\ \rho&1\end{pmatrix}$, denoting the points so generated by $\left(\tilde b_1^{(i)},\tilde b_2^{(i)}\right)$ for $i=1,...,M$. 
We then set $b_1^{(i)} = \sigma \tilde b_1^{(i)} + \mu_1$ and $b_2^{(i)} = \sigma \tilde b_2^{(i)}+ \mu_2$.

@@

\begin{equation}
\bar r_1^\#=\EX(\ln [(1-\delta)+\delta \exp (b_1^\#-b_2)]),  \label{rbar1sharp_E}
\end{equation}

@@

\begin{equation}
\EX(\ln (\frac{N_2(t+1)}{N_2(t)}))=0, \label{r2_is_zero}
\end{equation}

@@

\begin{equation}
\Delta I=\left(\bar r_1 - \bar r_1^\#\right)-\left(\bar r_2- \bar r_2^\#\right)=\bar r_1 - \bar r_1^\#+\bar r_2^\#  \label{DeltaI}
\end{equation}

\noindent has been defined in terms of expected values of elementary expressions of the random variables $\left(b_1,b_2\right)$, $\left(b_1^\#,b_2\right)$, and $\left(b_2^\#, b_2\right)$.

Having already specified that $b_1$ (and therefore also $b_1^\#$) is distributed as $N\left(\mu_1, \sigma^2\right)$, and that $b_2$ (and therefore also $b_2^\#$) is distributed as $N\left(\mu_2, \sigma^2\right)$, we know, using $\Sigma$ as notation for a covariance matrix that

\begin{equation}
\left(b_1^\#, b_2\right) \sim N\left((\mu_1,\mu_2),\Sigma \right) \label{b1sharpb2_distribution}
\end{equation}

@@

\begin{equation}
\left(b_2^\#, b_2\right) \sim N(( \mu_2,\mu_2), \Sigma) \label{b2sharpb2_distribution}
\end{equation}

@@

\begin{equation}
(b_1,b_2) \sim  N(( \mu_1,\mu_2), \Sigma) \label{b1b2_distribution_sym}
\end{equation}

@@

\begin{equation}
\Sigma = \begin{pmatrix} \sigma^2&\rho\\\rho&\sigma^2\end{pmatrix} \label{Sigmamat_rho}
\end{equation}

@@

\begin{enumerate}
\item Generate left-tail assocated noise $\left(\tilde b_{l1}^{(i)}, \tilde b_{l2}^{(i)}\right)$, $i=1,...,M$, as described in the section \ref{sect:noise}. Hence $\tilde b_{l1}^{i)}$ and $\tilde b_{l2}^{(i)}$ are distributed as $N(0,1)$. 
\item Generate right-tail associated noise $\left(\tilde b_{r1}^{(i)}, \tilde b_{r2}^{(i)}\right)$, $i=1,...,M$, via the analogous procedure. Again, $\tilde b_{r1}^{i)}$ and $\tilde b_{r2}^{(i)}$ are distributed as $N(0,1)$. 
\item Generate $M$ points $\tilde u_i$, $i = 1,...,M$ from a standard normal distribution. 
\end{enumerate}

@@

\item $\hat {\Delta I} = \hat \bar r_1 - \hat \bar r_1^{\#} + \hat \bar r_2^{\#}$, estimated separately for the left-tail associated, right-tail associated and symmetric noise cases. We obtained conservative standard error estimates by adding the standard error of the estimates of $\hat \bar r_1, \hat \bar r_1^{\#}$, and $\hat \bar r_2^{\#}$.

@@

Not only does the expected value $\bar r_1$ influence the capacity for species 1 to invade where rare, but so do other aspects of the distribution of
\begin{equation}
r_1 = \ln[1-\delta+\delta \exp(b_1-b_2)]. \label{r1_distribution}
\end{equation}

@@

We also considered the probability $\Prob[r_1 >0]$, which is, explicitly, the probability that environmental variables that permit invasion. We have
\begin{align}
\Prob[r_1>0] &= \Prob [1-\delta+\delta \exp(b_1-b_2) > 1] \\
&= \Prob[\exp (b_1-b_2)>1] \\
&= \Prob[b_1>b_2],
\end{align}
